\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-01-21 paper.qmd
Do the chnages in the introduction

To create low-dimensional representations (typically in 2D) of high-D data, it is common to apply dimension reduction (DR) techniques. Approaches for DR involve linear methods such as principal component analysis (PCA) [@Karl1901], non-linear methods such as multi-dimensional scaling (MDS) [@Torgerson1967]. In the past decade, many new non-linear dimension reduction (NLDR) techniques have emerged, such as t-distributed stochastic neighbor embedding (tSNE) [@Laurens2008] and uniform manifold approximation and projection (UMAP) [@Leland2018], designed to capture the complex and non-linear relationships present within high-D data [@Johnstone2009]. NLDR techniques are the 2D models of high-D data in our context.

Visualization of NLDR techniques (m-vis) [@article59] for the same high-D data is particularly important in understanding and finding the best representation. If we have done so, the 2D models can be quite different from the data structure shown in high-D space. Therefore, visualizing the 2D model in high-D space (m-in-ds) is more useful to answer different types of questions:

- Is there a best representation of high-D data or are they all providing  equivalent information? Is there a best parameter choice to fit the 2D model? How does the model change when it's parameters change?

@@

If we can't easily ask and answer these questions, the ability of understanding the models are restricted. To find the best 2D model and the parameter choices, better understanding of the underlying science is really important.

@@

Consider the high-dimensional data a rectangular matrix $X$, where $X = \begin{bmatrix} \textbf{x}_{1} & \textbf{x}_{2} & \cdots & \textbf{x}_{n}\\ \end{bmatrix}^\top$, with $n$ observations in $p$ dimensions. The objective is to discover a low-dimensional projection $Y = \begin{bmatrix} \textbf{y}_{1} & \textbf{y}_{2} & \cdots & \textbf{y}_{n}\\ \end{bmatrix}^\top$, represented as an $n$ × $d$ matrix, where $d \ll p$. The reduction process seeks to remove noise from the original data set while retaining essential information.

@@

#### Model selection {#sec-selection}

@@

Different NLDR techniques perform differently on the S-curve data and tSNE provides the best representation in 2D according to AIC and MSE values (see @fig-diagnosticpltScurve).

@@

@@

@@

## To find the benchmark value
benchmark_pbmc <- find_benchmark_value(.data = distance_pbmc, distance_col = distance)

@@

trimesh_removed_pbmc_umap <- trimesh_removed_pbmc_umap +
  geom_point(size = 2, colour = "#33a02c") +
  #ggtitle("(b)") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 30, min_dist = 0.3) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<https://youtu.be/lIA6ZPZAcLE>)"
#| label: fig-nldervisPBMCUMAP
 #| out-width: 100%

>>>>

To create low-dimensional representations (typically in 2D) (m-vis) [@article59] of high-D data, it is common to apply dimension reduction (DR) techniques. Approaches for DR involve linear methods such as principal component analysis (PCA) [@Karl1901], non-linear methods such as multi-dimensional scaling (MDS) [@Torgerson1967]. In the past decade, many new non-linear dimension reduction (NLDR) techniques have emerged, such as t-distributed stochastic neighbor embedding (tSNE) [@Laurens2008] and uniform manifold approximation and projection (UMAP) [@Leland2018], designed to capture the complex and non-linear relationships present within high-D data [@Johnstone2009]. NLDR techniques are the 2D models of high-D data in our context.

It is important to visualize various Non-Linear Dimensionality Reduction (NLDR) techniques for the same high-D data in order to understand and find the best representation. After doing so, the 2D models may differ considerably from each other and may also deviate from the original data structure in high-dimensional space. Therefore, visualizing the 2D model in high-D space (m-in-ds) is more useful to answer different types of questions:

- Is there a best 2D representation of high-D data or are they all providing  equivalent information? Is there a best parameter choice to fit the 2D model? How does the model change when it's parameters change?

@@

If we cannot easily ask and answer these questions, our ability to understand the models is limited. To find the best 2D model and parameter choices, a better understanding of the underlying science is important.

@@

Consider the high-dimensional data a rectangular matrix $X_{n \times p}$, where $X_{n \times p} = \begin{bmatrix} \textbf{x}_{1} & \textbf{x}_{2} & \cdots & \textbf{x}_{n}\\ \end{bmatrix}^\top$, with $n$ observations in $p$ dimensions. The objective is to discover a low-dimensional projection $Y_{n \times d} = \begin{bmatrix} \textbf{y}_{1} & \textbf{y}_{2} & \cdots & \textbf{y}_{n}\\ \end{bmatrix}^\top$, represented as an $n$ × $d$ matrix, where $d \ll p$. The reduction process seeks to remove noise from the original data set while retaining essential information.

@@

@@

<!--Different NLDR techniques perform differently on the S-curve data and tSNE provides the best representation in 2D according to AIC and MSE values (see @fig-diagnosticpltScurve). -->

@@

### Model selection {#sec-selection}

@@

- 15 PCs

@@

## To find the benchmark value
benchmark_pbmc <- find_benchmark_value(.data = distance_pbmc, distance_col = distance)
benchmark_pbmc <- 1.072513

@@

trimesh_removed_pbmc_umap <- trimesh_removed_pbmc_umap +
  geom_point(size = 1, colour = "#33a02c") +
  #ggtitle("(b)") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 30, min_dist = 0.3) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<https://youtu.be/CNXjWZMyh_M>)"
#| label: fig-nldervisPBMCUMAP
 #| out-width: 100%

### 2024-01-21 paper.tex
Do the chnages in the introduction

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, frame hidden, breakable, sharp corners, enhanced]}{\end{tcolorbox}}\fi

@@

To create low-dimensional representations (typically in 2D) of high-D
data, it is common to apply dimension reduction (DR) techniques.
Approaches for DR involve linear methods such as principal component
analysis (PCA) \citep{Karl1901}, non-linear methods such as
multi-dimensional scaling (MDS) \citep{Torgerson1967}. In the past
decade, many new non-linear dimension reduction (NLDR) techniques have
emerged, such as t-distributed stochastic neighbor embedding (tSNE)
\citep{Laurens2008} and uniform manifold approximation and projection
(UMAP) \citep{Leland2018}, designed to capture the complex and
non-linear relationships present within high-D data
\citep{Johnstone2009}. NLDR techniques are the 2D models of high-D data
in our context.

Visualization of NLDR techniques (m-vis) \citep{article59} for the same
high-D data is particularly important in understanding and finding the
best representation. If we have done so, the 2D models can be quite
different from the data structure shown in high-D space. Therefore,
visualizing the 2D model in high-D space (m-in-ds) is more useful to
answer different types of questions:

\begin{itemize}
\item
  Is there a best representation of high-D data or are they all
  providing equivalent information? Is there a best parameter choice to
  fit the 2D model? How does the model change when it's parameters
  change?
\item
  How well the does the 2D models capture the data structure? Is the
  model fitting able to capture different data structure like
  non-linear, clustering?
\end{itemize}

If we can't easily ask and answer these questions, the ability of
understanding the models are restricted. To find the best 2D model and
the parameter choices, better understanding of the underlying science is
really important.

@@

Consider the high-dimensional data a rectangular matrix \(X\), where
\(X = \begin{bmatrix} \textbf{x}_{1} & \textbf{x}_{2} & \cdots & \textbf{x}_{n}\\ \end{bmatrix}^\top\),
with \(n\) observations in \(p\) dimensions. The objective is to
discover a low-dimensional projection
\(Y = \begin{bmatrix} \textbf{y}_{1} & \textbf{y}_{2} & \cdots & \textbf{y}_{n}\\ \end{bmatrix}^\top\),
represented as an \(n\) × \(d\) matrix, where \(d \ll p\). The reduction
process seeks to remove noise from the original data set while retaining
essential information.

@@

\hypertarget{sec-selection}{%
\subsubsection{Model selection}\label{sec-selection}}

@@

Different NLDR techniques perform differently on the S-curve data and
tSNE provides the best representation in 2D according to AIC and MSE
values (see Figure~\ref{fig-diagnosticpltScurve}).

@@

@@

\begin{itemize}
\item
  Intro to the data set (This data set contains 13,714 features across
  2,700 samples within a single assay. The active assay is RNA, with
  13,714 features representing different gene expressions.)
\item
  2622 cells
\item
  Evaluate Festem on DEG detection (Festem enabled identification of
  often-missed fine cell types)
\item
  Using the Festem-selected genes, identified 10 clusters in the PBMCsk
  data and annotated them based on the expression of canonical markers
\item
  The 10 clusters included immune cells such as naive CD4 T cells,
  memory CD4 T cells, CD8 T cells, CD14 monocytes, FCGR3A monocytes,
  Natural Killer (NK) cells, B cells and Dendritic Cells (DC).
\item
  In addition to these common cell types, Festem identified a fine cell
  type, CD27− CD4+ memory T cells, which were often missed by other
  methods.
\item
  The CD27. CD4+ memory T cells identified by Festem expressed common
  marker genes (IL7R and S100A4) of memory T cells, but did not express
  CD27. These cells also had downregulated expression of SELL, CCR7, MAL
  and LEF1, and upregulated expression of CCL5 (Fig. 4B) and thus were
  the CD27- CD4+ memory T cells in the literature (36). The CD27- CD4+
  memory T cells were known to be at a more differentiated state and
  have stronger antigen-recall responses than their CD27+ counterparts.
\item
  learned from the model: There are 3 well-separated clusters in 2D, but
  if look at the model in high-D space, the three clusters are much
  closer. Also, there is some continuity within the clustering structure
  and it didn't capture well. Therefore UMAP with n:neighbour 30 is not
  a good representation for this data set.
\end{itemize}

@@

\caption{\label{fig-nldervisPBMCUMAP}(a) 2D layout from UMAP
(n\_neighbors = 30, min\_dist = 0.3) applied for the PBMC3k dataset. Is
this a best representation of the original data?, (b) Model in the 2D
space with UMAP (\url{https://youtu.be/lIA6ZPZAcLE})}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

@@

To create low-dimensional representations (typically in 2D) (m-vis)
\citep{article59} of high-D data, it is common to apply dimension
reduction (DR) techniques. Approaches for DR involve linear methods such
as principal component analysis (PCA) \citep{Karl1901}, non-linear
methods such as multi-dimensional scaling (MDS) \citep{Torgerson1967}.
In the past decade, many new non-linear dimension reduction (NLDR)
techniques have emerged, such as t-distributed stochastic neighbor
embedding (tSNE) \citep{Laurens2008} and uniform manifold approximation
and projection (UMAP) \citep{Leland2018}, designed to capture the
complex and non-linear relationships present within high-D data
\citep{Johnstone2009}. NLDR techniques are the 2D models of high-D data
in our context.

It is important to visualize various Non-Linear Dimensionality Reduction
(NLDR) techniques for the same high-D data in order to understand and
find the best representation. After doing so, the 2D models may differ
considerably from each other and may also deviate from the original data
structure in high-dimensional space. Therefore, visualizing the 2D model
in high-D space (m-in-ds) is more useful to answer different types of
questions:

\begin{itemize}
\item
  Is there a best 2D representation of high-D data or are they all
  providing equivalent information? Is there a best parameter choice to
  fit the 2D model? How does the model change when it's parameters
  change?
\item
  How well the does the 2D models capture the data structure? Is the
  model fitting able to capture different data structure like
  non-linear, clustering?
\end{itemize}

If we cannot easily ask and answer these questions, our ability to
understand the models is limited. To find the best 2D model and
parameter choices, a better understanding of the underlying science is
important.

@@

Consider the high-dimensional data a rectangular matrix
\(X_{n \times p}\), where
\(X_{n \times p} = \begin{bmatrix} \textbf{x}_{1} & \textbf{x}_{2} & \cdots & \textbf{x}_{n}\\ \end{bmatrix}^\top\),
with \(n\) observations in \(p\) dimensions. The objective is to
discover a low-dimensional projection
\(Y_{n \times d} = \begin{bmatrix} \textbf{y}_{1} & \textbf{y}_{2} & \cdots & \textbf{y}_{n}\\ \end{bmatrix}^\top\),
represented as an \(n\) × \(d\) matrix, where \(d \ll p\). The reduction
process seeks to remove noise from the original data set while retaining
essential information.

@@

@@

@@

\hypertarget{sec-selection}{%
\subsection{Model selection}\label{sec-selection}}

@@

\begin{itemize}
\item
  Intro to the data set (This data set contains 13,714 features across
  2,700 samples within a single assay. The active assay is RNA, with
  13,714 features representing different gene expressions.)
\item
  2622 cells
\item
  Evaluate Festem on DEG detection (Festem enabled identification of
  often-missed fine cell types)
\item
  Using the Festem-selected genes, identified 10 clusters in the PBMCsk
  data and annotated them based on the expression of canonical markers
\item
  The 10 clusters included immune cells such as naive CD4 T cells,
  memory CD4 T cells, CD8 T cells, CD14 monocytes, FCGR3A monocytes,
  Natural Killer (NK) cells, B cells and Dendritic Cells (DC).
\item
  In addition to these common cell types, Festem identified a fine cell
  type, CD27− CD4+ memory T cells, which were often missed by other
  methods.
\item
  The CD27. CD4+ memory T cells identified by Festem expressed common
  marker genes (IL7R and S100A4) of memory T cells, but did not express
  CD27. These cells also had downregulated expression of SELL, CCR7, MAL
  and LEF1, and upregulated expression of CCL5 (Fig. 4B) and thus were
  the CD27- CD4+ memory T cells in the literature (36). The CD27- CD4+
  memory T cells were known to be at a more differentiated state and
  have stronger antigen-recall responses than their CD27+ counterparts.
\item
  15 PCs
\item
  learned from the model: There are 3 well-separated clusters in 2D, but
  if look at the model in high-D space, the three clusters are much
  closer. Also, there is some continuity within the clustering structure
  and it didn't capture well. Therefore UMAP with n:neighbour 30 is not
  a good representation for this data set.
\end{itemize}

@@

\caption{\label{fig-nldervisPBMCUMAP}(a) 2D layout from UMAP
(n\_neighbors = 30, min\_dist = 0.3) applied for the PBMC3k dataset. Is
this a best representation of the original data?, (b) Model in the 2D
space with UMAP (\url{https://youtu.be/CNXjWZMyh_M})}
